NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-11-01 Manuscript/manuscriptPGPE.tex
Update text

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
\todo{There may be some relevant publication missing from here},
but glycerol backbone and headgroup structures have not
been evaluated against experiments.

@@

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to PE by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes (Fig.~\ref{HGorderparametersPCvsPEPG}).
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this?}.\\
\todo{Maybe we should analyze the P-N vector angle from different simulations?}.

@@

\todo{This is text by P. Fuchs, copied from the blog. \\
Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
- pure POPC \\
CHARMM36: 0.624 \\
Berger : 0.649 \\
- POPC/POPE 50:50 \\
CHARMM36 : POPC 0.609, POPE 0.557 \\
Berger-hacked: POPC 0.637, POPE 0.632 \\
----- \\
One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
In contrast, the values for Berger are not that changed. 
The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
-------\\
The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
which supports the more condensed PE bilayer.
This is interesting, but to avoid the overexapansion of the manuscript, it is probably better to keep the focus in headgroups and ion binding also in this manuscript.}

>>>>

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative
bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,henin09,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
\todo{There may be some relevant publication missing from here},
but evaluation of glycerol backbone and headgroup structures against experiments is rare \cite{henin09}.

@@

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to PE by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes (Fig.~\ref{HGorderparametersPCvsPEPG}).
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
\todo{This is text by P. Fuchs, copied from the blog. \\
Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
- pure POPC \\
CHARMM36: 0.624 \\
Berger : 0.649 \\
- POPC/POPE 50:50 \\
CHARMM36 : POPC 0.609, POPE 0.557 \\
Berger-hacked: POPC 0.637, POPE 0.632 \\
----- \\
One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
In contrast, the values for Berger are not that changed. 
The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
-------\\
The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
which supports the more condensed PE bilayer.
This is interesting, but not exactly the core message of the manuscript.
Maybe we should mention this very briefly? For example, we could just report the areas per lipid (without distinguishing PC and PE)
and mention the difference between CHARMM36 and Berger. I have opened an issue for this: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/7
}

None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this?}.\\
\todo{Maybe we should analyze the P-N vector angle from different simulations?}.

@@
